Appendix A: Forecast Trap in a multi-species model

### 2022-02-25 notebook/appendix-A.Rmd
appendix A

---
output: pdf_document
title: Appendix A: Multispecies models
---

The model which gives the best forecast does not always lead to the best decision.

We consider a simulated system described by the 5-dimensional model of a herring fishery presented in @Brias2021:

\begin{align}
\end{align}

@@

While richer numerical models can successfully capture the full complexity of this five-dimensional model, we must remember that reality is always more complex than our models.
As the term implies, models intentional simplifications of reality.

- Model A: A reasonable approximation model that produces more accurate forecasts but leads to substantially worse management outcomes
- Model B: A reasonable approximation model that does not produce very accurate forecasts, but nevertheless leads to good management decisions.
- Model C: The "True" Model: five-species model; functional forms and parameter values are never disclosed to the 'manager' but are used to simulate all data and management scenarios.

- Define a model
- Define action space. Under the assumption of fixed action, problem is not MDP and trivial to solve by sweep through possible fixed actions.

@@

```{r}
## helper funs
format_sim <- function(df) {
  df %>%  
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
  pivot_longer(any_of(c("cormorant", "bass", "herring")),
                  names_to = "species",
                  values_to = "biomass")
}
```

## Five species

@@

```{r utility}
clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}

## Compute the reward associated with a given simulation df given 
## rates for prey and predator species (x[1],x[2])
reward <- function(df,
                   x,
                   w3 = 0.5,              # Conservation value
                   w2 = 0.25,             # Predator harvest value
                   w1 = 1 - (w2+w3),      # Prey harvest value
                   delta = 0.999
                  ) {
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    R <-
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    as.numeric(R %*% delta ^ t)
}

# Function to be optimized for fixed choice of harvest rates x[1],x[2]
utility <- function(x, f = model_5sp, ...) {

o <- optim(c(0.01,0.5), utility, method="L-BFGS-B", lower = c(0,0), upper = c(1,1))
o$par
o$value

o <- optim(c(0.01,0.5), utility)
clip(o$par, 0, 1)
o$value

x <- c(0.0,.58)
utility(x)

For the sake of argument, let us allow the model of Brias & Munch to represent the "true" model, which is unknown.
As researchers, we only have access to data derived from observations from the model. 
Let us consider a sample simulation of such data:

There are many other models, including process-based models, statistical models or even black-box machine learning models
which 5 dimensional state-space and two-dimensional action space.

Now consider an alternative model which ignores some of this biological complexity:
Our alternate model has several oversimplifications:
(1) it treat all three herring species as a single species,
(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
and ignoring the impact of cormorant's predation on the herring itself.
(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.

The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the unfished dynamics.
Under any harvesting regime or other influence that perturbs the system significantly far from the unfished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.

```{r}
model_3sp <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 1.6,
                      A2 = 2.75,
                      R2 = 1.45,
                      Tmax = 50,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      sigma = 0.01,
                      reps=40) {
    df <- tibble()
    for(r in 1:reps){
    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

@@

```{r}
# Optimal policy under identical utility
u_3sp <- function(x) utility(x, f = model_3sp)

m3sp <- optim(c(0.5,0.1), u_3sp)
m3sp$value
clip(m3sp$par, 0, 1)

@@

Alternate model, also mistakenly treats the three species of herring as a single species, but does a better job capturing the predator-prey relationships.  This model also happens to produce better forecasts, but leads to worse decision:

```{r}
model_B <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 =  3, #1.15, #1.0213,
                      A11 = 8,
                      A14 = 0.9, # 0.7252,
                      A15 = 0.9, # 0.2445,
                      A41 =  0.1, #0.7252,
                      A51 = 0.3, # 0.2445,
                    reps = 40) {

```{r}
# Optimal policy under identical utility
u_B <- function(x) utility(x, f = model_B)
u_B(c(0, .58))
#optim(c(0.5,0.1), u_B, method = "L-BFGS-B", lower=0, upper=1)

B <- optim(c(.2,0.01), u_B)
model_B_policy <- clip(B$par, 0, 1)
model_B_policy

Run the simulation using management policy of each

Tmax <- 20
x <- model_B_policy
#x <- c(0.,0.58)  # Following the optimal policy under model_B

sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma= 0.05)
obs <- sim %>% format_sim()
predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

reward(df = sim, x = x) / 40
```

```{r}
obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict_stats_B <- predict_B %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_B")

bind_rows(obs_stats, predict_stats_B) %>% 
  filter(species != "herring") %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_point(aes(shape=type)) +
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))

```{r}
## True observations vs forecasts, 5 steps ahead, while following the Optimal / Model 1 policy, starting with default initial condition
Tmax <- 10
x <- c(0,0.58)

sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax, sigma=0.05)
obs <- sim %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

reward(df = sim, x = x) / 40
```

```{r}
obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict_stats <- predict %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted_A")

bind_rows(obs_stats, predict_stats) %>% 
  filter(species != "herring") %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_point(aes(shape=type)) +
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type))
```

```{r}
score_me <- function(x = c(0,0.58), process=model_5sp, forecast=model3p, Tmax=5) {

  obs <- process(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
  predict <- forecast(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

}
```

## unfished

```{r}
## True observations vs forecasts, 5 steps ahead, unfished, starting with default initial condition
Tmax <- 20
x <- c(0,0.)

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], sigma=0.05, Tmax = Tmax) %>% format_sim()
predict <- model_3sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats <- predict %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

## dotted are forecasted values, solid are observed values

predict_stats %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))

Tmax <- 5
x <- c(0,0)
x <- model_B_policy
#x <- c(0.,0.58)  # Following the optimal policy under model_B

obs <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()
predict_B <- model_B(u_prey = x[[1]], u_pred = x[[2]], Tmax = Tmax) %>% format_sim()

obs_stats <- obs %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats_B <- predict_B %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predict_stats_B %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd), lty=2) + 
  geom_errorbar(data = obs_stats, aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd))

Does this model produce good forecasts?

```{r}
# unfished equilibrium
m1_forecast <- model_3sp(0, 0.0, sigma = 0.05,  Tmax = 50)

predictions <- m1_forecast %>%
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
  pivot_longer(any_of(c("cormorant", "bass", "herring")),
                  names_to = "species",
                  values_to = "biomass") %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predictions %>%
  ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
  geom_point(data = filter(obs, r==1))
```

```{r}
# unfished equilibrium
m1_forecast <- model_3sp(0, 0.58, sigma = 0.05,  Tmax = 50)

predictions <-   
m1_forecast %>%
    transmute(time = time, cormorant = x5, bass = x4, herring = x1, r=r) %>%
    pivot_longer(any_of(c("cormorant", "bass", "herring")),
                    names_to = "species",
                    values_to = "biomass") %>% 
    group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass))

predictions %>%
ggplot(aes(time, biomass, col= species)) + 
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd)) +
  geom_point(data = filter(obs, r==1))
```

>>>>

---
title: 'Appendix A: Forecast Trap in a multi-species model'
author: Carl Boettiger
journal: Ecology Letters
layout: 3p
bibliography: refs.bib
csl: csl/ecology-letters.csl
#output:
#  tufte::tufte_handout:
#    dev: cairo_pdf
#    latex_engine: xelatex
output: 
  hrbrthemes::ipsum_pdf:
    dev: cairo_pdf
    latex_engine: xelatex

header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers

---

This appendix illustrates how the forecast trap can emerge in choosing between two alternative three-species models which may be used to manage a partially observed five-species system. 
The underlying dynamics for the simulated "true" system are based on the five-dimensional model of a herring fishery presented in @Brias2021.
Following @Brias2021, we assume three species of herring, $X_1$, $X_2$, $X_3$ are preyed upon by

\begin{align}
x_{1,t+1} &= s_{1,t} \exp{(1.0213 -        s_{1,t} - 0.0861 s_{2,t} - 0.3141 s_{3,t} - 0.7252 s_{4,t} -0.2445 x_{5,t} +\espilon_{1,t} )}
x_{2,t+1} &= s_{2,t} \exp{(1.0289 - 0.4765 s_{1,t} -        s_{2,t} - 0.1370 s_{3,t} - 0.9811 s_{4,t} -0.0915 x_{5,t} +\espilon_{2,t} )}
x_{3,t+1} &= s_{3,t} \exp{(1.0207 - 0.3193 s_{1,t} - 0.3461 s_{2,t} -        s_{3,t} - 0.6367 s_{4,t} -0.8716 x_{5,t} +\espilon_{3,t} )}
x_{4,t+1} &= s_{4,t} \exp{(         0.7252 s_{1,t} + 0.9811 s_{2,t} + 0.6367 s_{3,t} -        s_{4,t}                 +\espilon_{4,t})}
x_{5,t+1} &= x_{5,t} \exp{(         0.2445 s_{1,t} + 0.0915 s {2,t} + 0.8716 s {3,t} -                        x_{5,t} +\espilon_{5,t} )}
\end{align}

where

\begin{equation}
s_{i,t} = x_{i_t} \left(1 - u_{i,t}\right)
\end{equation}

\begin{equation}
\epsilon_{i,t} \stackrel{\texrm{iid}}{\tilde} \mathcal{N(0, \sigma_i)}
\end{equation}

@@

Herring are harvested together, $u_1 = u_2 = u_3 = u_{\textrm prey}$, while bass are harvested at effort $u_{\textrm predator}$.
The manager's utility in year $t$ is given by

\begin{equation}
R(\vec x_t, \vec u_t) = w_3 x_5 + w_2 x_4 + \sum_i^{1,2,3} w_1 x_i
\end{equation}

We will consider the case which places 50% of the weight on the cormorant conservation objective, $w_3 = 0.5$, and 25% weight on the harvest of bass and herring respectively, $w_2 = w_1 = 0.25$.

In contrast to @Brias2021, we will assume that the manager must simply select the harvest efforts $u_{\textrm prey} \in [0,1]$ and $u_{\textrm predator} \in [0,1]$ to be used over the duration of the management scenario, rather than being able to adjust these annually.
In most real-world systems it is not possible to measure all variables which influence the dynamics, some dimensions are treated as latent variables of the model.
To illustrate this, we will assume that the manager observes only the population abundance of bass $x_{4,t}$, and cormorants, $x_{5,t}$, but not the population size of any of the three species of herring.

### Model A

Model A is a possible 3-species approximation that produces more accurate forecasts but leads to substantially worse management outcomes
This model treats the three species of herring collectively as $x_1$. 
Parameters are estimated consistent with limited historical data of the observed variables, bass biomass $x_4$ and herring biomass $x_5$.

\begin{equation}
x_{1,t+1} &= s_{1,t} \exp(3 - 8 s_{1,t} - 0.9 s_4 - 0.9 * x_{5,t} + \espilon_{1,t} )
x_{4,t+1} &= s_{4,t} \exp(0.1 s_{1,t} - s_{4,t} + \espilon_{4,t}  )
x_{5,t+1} &= x_{5,t} \exp(0.3 s_{1,t} - x_5  + \espilon_{5,t}  )
\end{equation}

### Model B

Model B is a structurally different model which postulates a further simplification of the dynamics.
This slightly over-simplified approximation does not produce very accurate forecasts, but nevertheless leads to good management decisions.

\begin{equation}
x_{1,t+1} &= s_{1,t} \exp(1.60 - 2.00 s_{1,t}  + \espilon_{1,t} )
x_{4,t+1} &= s_{4,t} \exp(1.45 - 2.75 s_{1,t}  + \espilon_{4,t}  )
x_{5,t+1} &= 0.25 x_{5,t} 
\end{equation}

Our alternate model has several oversimplifications:
(1) it treat all three herring species as a single species,
(2) it fails to capture the coupled predator-prey dynamics between bass and herring, 
(3) it oversimplifies the cormorant dynamics, assuming the cormorant population is determined to be a fixed fraction of the herring,
and ignoring the impact of cormorant's predation on the herring itself.
(4) Lastly, our model will overestimate the mortality introduced on herring by a given fishing effort.
These elements are all obviously wrong, but not so arbitrary as to be inconceivable as a candidate model.
Researchers frequently consider models which make oversimplifications all the time, and rely on model choice processes to weed them out.

The parameterization chosen for the oversimplified 3 species model can reasonably reproduce the un-fished dynamics.
Under any harvesting regime or other influence that perturbs the system significantly far from the un-fished equilibrium co-existence state of the model quickly reveals the poor forecasting ability of this model.
Nevertheless, solving for the optimal policy (under the same constraints of constant harvest fractions as we consider for the true model) results in a harvest policy which provides nearly optimal performance.

# Software implementation

```{r graphics_setup, message=FALSE, warning=FALSE, include = FALSE}
## Plotting themes, colors, and fonts 
## aesthetics choices only, all can be omitted without consequence
library(ggthemes)
library(hrbrthemes)
library(ggplot2)
library(Cairo)
library(extrafont)
library(patchwork)
library(styler)
extrafont::loadfonts(quiet = TRUE)
ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())

scale_colour_discrete <- function(...) ggthemes::scale_colour_solarized()
scale_fill_discrete <- function(...) ggthemes::scale_fill_solarized()
pal <- ggthemes::solarized_pal()(8)
txtcolor <- "#586e75"

knitr::opts_chunk$set(cache=FALSE, tidy = "styler", message = FALSE, warning = FALSE, echo = TRUE)

@@

```{r}
## helper funs
format_sim <- function(df) {
  df %>%  
  transmute(time = time, cormorant = x5, bass = x4, herring = x1+x2+x3, r=r) %>%
  pivot_longer(any_of(c("cormorant", "bass", "herring")),
                  names_to = "species",
                  values_to = "biomass")
}

clip <- function(x, lower=0, upper=1) {
    x[x <= lower] <- lower
    x[x >= upper] <- upper
    x
}
```

Define the reward for a given state,action pair and the net utility

```{r utility}
## Compute the reward associated with a given simulation df given 
## rates for prey and predator species (x[1],x[2])
reward <- function(df,
                   x,
                   w3 = 0.5,              # Conservation value
                   w2 = 0.25,             # Predator harvest value
                   w1 = 1 - (w2+w3),      # Prey harvest value
                   delta = 0.999
                  ) {
    u_prey <- clip(x[1], 0, 1)
    u_pred <- clip(x[2], 0, 1)
    R <-
        w1 * u_prey * (df$x1 + df$x2 + df$x3) +
        w2 * u_pred * df$x4 +
        w3 * df$x5
    t <- seq_along(R)
    as.numeric(R %*% delta ^ t)
}

# Function to be optimized for fixed choice of harvest rates x[1],x[2]
utility <- function(x, f, ...) {

## Model definitions

### True five species-model

@@

### Model A

```{r}
model_A <- function(u_prey = 0.5,
                      u_pred = 0.5,
                      sigma = 0.05,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      Tmax = 50,
                      R1 =  3, #1.15, #1.0213,
                      A11 = 8,
                      A14 = 0.9, # 0.7252,
                      A15 = 0.9, # 0.2445,
                      A41 =  0.1, #0.7252,
                      A51 = 0.3, # 0.2445,
                    reps = 40) {

### Model B

```{r}
model_B <- function(u_prey = 0,
                      u_predator = 0,
                      A1 = 2,
                      R1 = 1.6,
                      A2 = 2.75,
                      R2 = 1.45,
                      Tmax = 50,
                      x1o = 0.44, # unfished equib
                      x2o = 0.17,
                      x3o = 0.186,
                      x4o = 0.600,
                      x5o = 0.280,
                      sigma = 0.01,
                      reps=40) {
    df <- tibble()
    for(r in 1:reps){
    omega1 <- rnorm(Tmax, 0, sigma)
    omega2 <- rnorm(Tmax, 0, sigma)
    omega3 <- rnorm(Tmax, 0, sigma)

@@

## Optimal decision

Determine the optimal policy using Nelder-Meade simplex method provided by the `optim` function:

```{r}
u_A <- function(x) utility(x, f = model_A)
A <- optim(c(.2,0.01), u_A)
model_A_policy <- clip(A$par, 0, 1)
```

```{r}
u_B <- function(x) utility(x, f = model_B)
B <- optim(c(0.5,0.1), u_B)
model_B_policy <- clip(B$par, 0, 1)
```

```{r}
u_true <- function(x) utility(x, f = model_5sp)
o <- optim(c(0.01,0.5), u_true)
true_policy <- clip(o$par, 0, 1)
```

## Simulate scenarios

We simulate the system using the true model, while managing from the policy derived from forecasts of model A:

```{r}
Tmax <- 10
resp <- 40

x <- model_A_policy
sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]],
                 Tmax = Tmax, sigma= 0.05, reps = reps)
obs <- sim %>% 
  format_sim() %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict <- model_A(u_prey = x[[1]], u_pred = x[[2]],
                     Tmax = Tmax, reps = reps) %>%
  format_sim() %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted")

scenario_A <- bind_rows(obs, predict) %>% 
  filter(species != "herring") %>% mutate(scenario = "Model A")
mean_utility_A <- reward(df = sim, x = x) / reps
```

We repeat the simulation over 40 replicates using the policy derived from model B forecasts:

```{r}
x <- model_B_policy
sim <- model_5sp(u_prey = x[[1]], u_pred = x[[2]], 
                 Tmax = Tmax, sigma=0.05, reps = reps)
obs <- sim %>%
  format_sim() %>%
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "observed")

predict <- model_B(u_prey = x[[1]], u_pred = x[[2]], 
                   Tmax = Tmax, reps = reps) %>% 
  format_sim()  %>% 
  group_by(time, species) %>% 
  summarize(sd = sd(biomass), biomass = mean(biomass), type = "predicted")

scenario_B  <- bind_rows(obs, predict) %>% 
  filter(species != "herring")  %>% mutate(scenario = "Model B")

mean_utility_B <- reward(df = sim, x = x) / reps
```

example1 <- bind_rows(scenario_A, scenario_B)
write_csv(example1, "../data/example1.csv")

example1 %>% 
  ggplot(aes(time, biomass, col= species)) + 
  geom_point(aes(shape=type)) +
  geom_errorbar(aes(ymin = biomass - 2*sd, ymax = biomass + 2 *sd, lty=type)) +
  facet_wrap(~scenario)

@@

```{r}
mean_utility_A / mean_utility_B
```
